---
title: "Introduction to HoopsImpact"
author: "Robbie"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to HoopsImpact}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

HoopsImpact is an R package that brings together a small but practical
set of functions for basketball analytics. The package was built for
students, analysts, and coaches who want to evaluate player
contributions without having to write modeling code from scratch. This
vignette walks through the main features of the package using the
bundled sample dataset.

```{r setup}
library(HoopsImpact)
data(sample_players)
head(sample_players)
```

# Calculating Impact

The first piece of the package is the `impact_score()` function, which
combines positive box-score contributions with a turnover penalty and
scales the result by minutes played so that lower-minute players are
not unfairly punished. The formula used is:

$$
\text{Impact} = (\text{PTS} + 1.2 \cdot \text{REB} + 1.5 \cdot \text{AST} + 2 \cdot \text{STL} + 2 \cdot \text{BLK} - \text{TO}) \cdot \frac{\text{MIN}}{36}
$$

```{r}
impact_score(sample_players)
```

The `scale = TRUE` argument standardizes the scores so that they can be
compared across datasets:

```{r}
impact_score(sample_players, scale = TRUE)
```

# Classifying Player Roles

The `classify_role()` function assigns each player a role label based
on a simple set of thresholds. The categories were chosen because they
map cleanly to how coaches and analysts already describe players.

```{r}
classify_role(sample_players)
```

The thresholds are configurable, which means a user can tighten or
loosen the definitions to fit a specific league or level of play.

```{r}
classify_role(sample_players, scorer_pts = 18, general_ast = 6)
```

# Comparing Two Players

The `compare_players()` function builds a side-by-side comparison
across the core box-score categories and flags the player with the
edge in each one. Turnovers are handled correctly: the smaller value
wins.

```{r}
compare_players(sample_players, "A. Carter", "B. Mitchell")
```

# Visualizing a Single Player

The `plot_player()` function returns a ggplot2 bar chart of a single
player's per-game contributions. The output is a regular ggplot
object, so it can be customized with additional layers if needed.

```{r, fig.width = 6, fig.height = 4}
plot_player(sample_players, "A. Carter")
```

# S3 Player Profiles

HoopsImpact uses both R class systems so that users can see how each
one is built and used. The `player_summary()` function returns a
`player_profile` S3 object with custom `print()` and `summary()`
methods.

```{r}
prof <- player_summary(sample_players, "A. Carter")
print(prof)
```

```{r}
summary(prof)
```

# S4 Team Impact Objects

For team-level work the package uses an S4 class called `TeamImpact`.
S4 was chosen here because team objects benefit from formal slot
typing and a validity check that prevents the player and impact
vectors from drifting out of sync.

```{r}
team <- PlayerProfile(sample_players,
                      team_name = "Sample U",
                      season = "2025-26")
team
```

The `summary()` method returns a ranked data frame so that the most
impactful players appear at the top:

```{r}
summary(team)
```

# Conclusion

HoopsImpact pulls together the three things a basketball analyst
usually wants to do with a player dataset: score the players, label
the players, and compare the players. The combination of S3 and S4
classes makes it a useful teaching package as well as a practical one,
and the bundled sample dataset means every function can be tested
immediately after installation.
